\documentclass[6pt,landscape,a4paper]{extarticle}
\usepackage[landscape,margin=0.4cm]{geometry}
\usepackage{multicol}
\usepackage{amsmath,amssymb}
\usepackage{enumitem}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{titlesec}
\usepackage{listings}

\pagestyle{empty}
\setlength{\parindent}{0pt}
\setlength{\parskip}{1.2pt}
\setlength{\columnsep}{6pt}

\titleformat{\section}{\bfseries\normalsize}{}{0em}{}[\titlerule]
\titleformat{\subsection}{\bfseries\small}{}{0em}{}
\titlespacing{\section}{0pt}{3pt}{2pt}
\titlespacing{\subsection}{0pt}{2pt}{1pt}

\lstset{basicstyle=\ttfamily\fontsize{4.5}{5.5}\selectfont, breaklines=true, frame=none, columns=fullflexible}

\begin{document}

% ================================================================
% PAGE 1 --- CONCEPTUAL NOTES
% ================================================================
\begin{multicols}{3}

\section{Foundational Regression \& Classification}

\textbf{Linear Regression vs.\ Logistic Regression:}
Linear regression is linear. Logistic regression is a sigmoid classifier.

\section{Loss Functions}

\textbf{Step Loss:} Bad to use because it is not differentiable.

\textbf{MSE (Mean Squared Error):} Doesn't penalize wrong classification answers well enough (does penalize outlier numerical answers decently).

\textbf{Sigmoid:} An activation function that squashes inputs into a range between 0 and 1, often used in binary classification. Prone to saturation, which causes vanishing gradients in deep networks.

\textbf{Cross-Entropy:} A log-based loss function that penalizes incorrect classifications harshly, with the cost approaching infinity as the predicted probability for the true class approaches zero.

\textbf{Cross-Entropy \& Softmax:} Softmax is the activation function that turns raw scores (logits) into a probability distribution summing to 1. Cross-Entropy is the loss function that measures the distance between that distribution and the one-hot encoded true label.

\section{Regularization Techniques}

\textbf{L1 Regularization (Lasso):} Adds the sum of absolute weights to the loss; it removes a lot of features, producing a simpler model.

\textbf{L2 Regularization (Weight Decay/Ridge):} Adds the sum of squared weights to the loss; spreads evenly across features, use it if you want to preserve features because you think they all have some significance.

\textbf{Elastic Net:} Combines L1 and L2 using separate constants.

\textbf{Dropout:} Randomly sets neurons to zero each forward pass. Forces redundant learning.

\textbf{Batch Normalization:} Shifts the distribution to a mean of zero and a variance of 1. Good for deep CNNs where data is regular and batches are large.

\section{Optimization Algorithms}

\textbf{Full-batch Gradient Descent:} Uses the entire dataset to compute one gradient update. Smooth loss and expensive.

\textbf{Mini-batch SGD:} Computes updates using a subset of the data, and updates based on a noisy estimate of the true gradient.

\textbf{SGD + Momentum:} SGD that has ``momentum'' that lets it move over some local minima.

\textbf{AdaGrad:} Adaptive algorithm that scales the learning rate for each parameter based on historical squared gradients.

\textbf{RMSProp:} Variation of AdaGrad that uses a moving average of squared gradients to prevent the learning rate from shrinking too quickly.

\textbf{Adam:} Combination of Momentum and RMSProp; often faster than standard SGD and handles sparse gradients well.

\textbf{AdamW:} Decouples weight decay from gradient updates in Adam.

\section{Multi-Layer Perceptrons (MLP)}

Simplest Neural Network. Very robust since everything is connected. Lots of sources means overfitting is a big risk.

\textbf{Feature Transforms:} Projecting data into a new geometric space where previously inseparable data becomes linearly separable.

\textbf{Non-linearity:} Activation functions like ReLU or Sigmoid stop it from being linear, letting it learn more.

\textbf{Forward Pass:} Applying the weights in order through the neural network to make a prediction.

\textbf{Backpropagation:} Applying the chain rule to compute the partial derivative of the loss with respect to every weight, allowing for model updates.

\section{Convolutional Neural Networks (CNN)}

\textbf{CNN vs.\ MLP:} CNNs use weight sharing and locality (receptive fields) to handle spatial data, they reuse neurons in a way that MLP does not, so they learn more on spatial data.

\textbf{Components:} Convolutional layers for feature extraction, Pooling layers (Max/Average) for downsampling, Fully Connected layers for final classification.

\textbf{Tensor Dimensions:}
$\text{Output} = \left\lfloor \frac{W - F + 2P}{S} \right\rfloor + 1$
\quad ($W$=input, $F$=filter, $P$=padding, $S$=stride)

\section{CNN Architectures}

\begin{itemize}[nosep,leftmargin=*]
  \item \textbf{AlexNet} (2012) --- ReLU over tanh, 6$\times$ faster training
  \item \textbf{VGG} (2014) --- 3$\times$3 filters, more depth (19 layers)
  \item \textbf{GoogLeNet} (2014) --- Inception modules with parallel paths
  \item \textbf{ResNet} (2015) --- Skip Connections to solve vanishing gradients in very deep (152-layer) networks
  \item \textbf{ConvNeXt} (2022) --- Outcompetes transformers sometimes
\end{itemize}

\section{Training Your Neural Network}

\textbf{Activation Functions:} ReLU is the default for CNNs; Leaky ReLU fixes ``Dying ReLU'' (neurons stuck at 0); Sigmoid is reserved for output layers in binary tasks. ReLU is good because it is ``jagged'' across time.

\textbf{Weight Initialization:} He initialization for ReLU networks, double the variance of Xavier to make up for ReLU jaggedness; Xavier for Sigmoid/Tanh to keep activation scales stable.

\textbf{Choosing Hyperparameters:} Key ``knobs'' include learning rate (most important), batch size, and weight decay (L2 regularization).

\section{RNN and LSTM}

\textbf{Language Modeling:} Predicting the next word or character in a sequence based on previous context.

\textbf{Sequence-to-Sequence:} An encoder-decoder framework for tasks like translation, where one RNN compresses the input and another generates the output.

\textbf{RNN vs.\ LSTM:} Standard RNNs struggle with vanishing gradients over long sequences. LSTMs solve this using a Cell State (long-term memory highway) and gates (Forget, Input, Output) to regulate information flow.

\end{multicols}

\newpage

% ================================================================
% PAGE 2 --- FORMULAS & REFERENCE TABLES
% ================================================================
\begin{multicols}{3}

\section{Loss Function Formulas}

\textbf{MSE}: $\frac{1}{P}\sum_{i=1}^{P}(\hat{y}_i - y_i)^2$

\textbf{Sigmoid}: $\sigma(t) = \frac{1}{1+e^{-t}}$

\textbf{Binary Cross-Entropy}: $-\frac{1}{P}\sum[y\log\hat{y} + (1\!-\!y)\log(1\!-\!\hat{y})]$

\textbf{Softmax}: $\text{softmax}(z_i) = \frac{e^{z_i}}{\sum_j e^{z_j}}$

\textbf{Multi-class CE}: $-\frac{1}{P}\sum_p \log(\text{softmax prob of true class})$

\textbf{Normal Equation}: $\mathbf{w} = (X^TX)^{-1}X^TY$

\subsection{Logistic vs Softmax at a Glance}
\begin{tabular}{@{}l|l|l@{}}
& \textbf{Logistic} & \textbf{Softmax} \\\hline
Classes & 2 only & Any $C\ge2$ \\
Output & Single prob & $C$ probabilities \\
Activation & Sigmoid & Softmax \\
Loss & Binary CE & Multi-class CE \\
\end{tabular}

\section{Regularization Formulas}

Total loss: $\text{data loss} + \lambda \cdot \text{penalty}$

\textbf{L2 (Ridge)}: $\lambda\sum W^2$

\textbf{L1 (Lasso)}: $\lambda\sum |W|$

\textbf{Elastic Net}: $\lambda_1\sum|W| + \lambda_2\sum W^2$

\textbf{Layer order}: Linear $\to$ BatchNorm $\to$ Activation $\to$ Dropout

\section{Optimizer Formulas}

\textbf{Core update}: $W \leftarrow W - \alpha \cdot (\text{direction})$

\textbf{Momentum}: $v_{t+1} = \rho v_t + \nabla L$, \quad $w \leftarrow w - \alpha v$

\textbf{AdaGrad}: $gs = gs + dW^2$, \quad $w \leftarrow w - \frac{\alpha}{\sqrt{gs}+\epsilon}\,dW$

\textbf{RMSProp}: $gs = \gamma\!\cdot\!gs + (1\!-\!\gamma)\!\cdot\!dW^2$, \quad $w \leftarrow w - \frac{\alpha}{\sqrt{gs}+\epsilon}\,dW$

\textbf{Adam}: 1st moment $m = \beta_1 m + (1\!-\!\beta_1)\nabla L$, 2nd moment $v = \beta_2 v + (1\!-\!\beta_2)(\nabla L)^2$, bias-corrected: $\hat{m}=\frac{m}{1-\beta_1^t}$, $\hat{v}=\frac{v}{1-\beta_2^t}$, update: $w \leftarrow w - \frac{\alpha\hat{m}}{\sqrt{\hat{v}}+\epsilon}$

Defaults: $\beta_1\!=\!0.9$, $\beta_2\!=\!0.999$, $\text{lr}\!=\!10^{-3}$

\textbf{AdamW}: $\theta \leftarrow \theta - \alpha\lambda\theta$ (decay separate from gradient)

\subsection{Optimizer Comparison}
\begin{tabular}{@{}l|cccc@{}}
& Mom. & Adaptive & Leaky & Bias corr \\\hline
SGD & & & & \\
+Momentum & \checkmark & & & \\
AdaGrad & & \checkmark & & \\
RMSProp & & \checkmark & \checkmark & \\
Adam & \checkmark & \checkmark & \checkmark & \checkmark \\
\end{tabular}

\section{Activation Functions}

\textbf{Sigmoid}: $\sigma(x)=\frac{1}{1+e^{-x}}$ \quad output $(0,1)$

\textbf{Tanh}: $\tanh(x)$ \quad output $(-1,1)$

\textbf{ReLU}: $\max(0,x)$

\textbf{Leaky ReLU}: $\max(0.01x,\, x)$

\textbf{GELU}: $x\,\Phi(x)$

\section{Weight Initialization}

\textbf{Xavier/Glorot} (sigmoid/tanh): $\text{std} = \sqrt{1/D_{in}}$

\textbf{Kaiming/He} (ReLU): $\text{std} = \sqrt{2/D_{in}}$

Conv layers: $D_{in} = C_{in} \times K \times K$

\section{CNN Dimension Formula}

\[\boxed{\text{Output} = \left\lfloor\frac{W + 2P - F}{S}\right\rfloor + 1}\]

CONV weights: $(F \!\times\! F \!\times\! C_{in}) \times C_{out} + C_{out}$ biases

Pool: same formula, depth unchanged, \textbf{0 params}

FC: flatten $\times$ neurons + biases

\subsection{Calculation Example}
INPUT $128\!\times\!128\!\times\!3$, stride 1, pad 0:
\begin{itemize}[nosep,leftmargin=*]
\item CONV-9-32: $120^2\!\times\!32$, wt$=9^2\!\cdot\!3\!\cdot\!32\!=\!7776$
\item POOL-2: $60^2\!\times\!32$, wt$=0$
\item CONV-5-64: $56^2\!\times\!64$, wt$=5^2\!\cdot\!32\!\cdot\!64\!=\!51200$
\item POOL-2: $28^2\!\times\!64$
\item CONV-5-64: $24^2\!\times\!64$, wt$=5^2\!\cdot\!64\!\cdot\!64\!=\!102400$
\item POOL-2: $12^2\!\times\!64$
\item FC-128: wt$=9216\!\cdot\!128$
\item FC-3: wt$=128\!\cdot\!3\!=\!384$
\end{itemize}

\section{RNN \& LSTM Formulas}

\textbf{RNN hidden state}: $h_t = \tanh(W_{hh}h_{t-1} + W_{xh}x_t + b)$

\textbf{RNN output}: $\hat{y}_t = \text{softmax}(W_y h_t)$

\textbf{LSTM Cell update}: $C_t = f \odot C_{t-1} + i \odot g$

\textbf{4 Gates} (all sigmoid $\to [0,1]$):
\begin{enumerate}[nosep,leftmargin=*]
\item Forget ($f$): erase old memory
\item Input ($i$): gate new info
\item Cell ($g$): candidate values (tanh)
\item Output ($o$): what to expose
\end{enumerate}

\textbf{Perplexity}: $\exp(\text{avg CE loss})$ --- lower is better

\textbf{Gradient clipping}: if $\|g\| > \tau$: $g \leftarrow g \cdot \tau/\|g\|$

\section{ResNet Skip Connection}

$H(x) = F(x) + x$

\section{PyTorch Essentials}
\begin{lstlisting}
optimizer.zero_grad()  # clear old grads
loss.backward()        # compute new grads
optimizer.step()       # update weights
\end{lstlisting}

\texttt{nn.CrossEntropyLoss}: input = raw logits (NOT softmax!)

\end{multicols}
\end{document}
